---
title: | 
  | \LARGE ASM Homework 1
subtitle: "Linear Model for IDMB data"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "13/10/2019"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE,results="hide"}

############################################################
###############  libraries  & Functions      ############### 
############################################################

### load packages ###
requireorinstall=function(package=""){
  reqpac=parse(text=paste("require(",as.character(package),")"))
  if(eval(reqpac)){
    print(paste(as.character(package), "has been loaded correctly"))
  } else {
    print(paste("trying to install" ,as.character(package)))
    eval(parse(text=paste("try(install.packages(",as.character(package),"))")))
    if(eval(reqpac)){
      print(paste(as.character(package) ,"has been installed and loaded correctly"))
    } else {
      warning(paste("could not install",as.character(package)))
    }
  }
}


requireorinstall(c("knitr","kableExtra","ggplot2","ggsci","dplyr","tidyr","lmtest",
                   "tidyverse","broom","purrr","magrittr","grid","gridExtra","scales",
                   "GGally","fBasics", "car", "effects", "emmeans","rockchalk"))

## theme definition for ggplot 
ggstyle = theme(panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_blank(),
                axis.line = element_blank(),
                axis.title.x = element_text(size = 10,
                                            vjust=0.1),
                axis.title.y = element_text(size = 10,
                                            vjust=1.5),
                axis.text.y = element_text(size=9),
                axis.text.x = element_text(size=9,
                                           angle = 45,
                                           vjust = 1,
                                           hjust=1),
                plot.title = element_text(size=11),
                legend.text = element_text( size=8),
                legend.title = element_text( size=9))

ggstyleFonts = theme(axis.title.x = element_text(size = 10,
                                                 vjust=0.1),
                     axis.title.y = element_text(size = 10,
                                                 vjust=1.5),
                     axis.text.y = element_text(size=9),
                     axis.text.x = element_text(size=9,
                                                angle = 45,
                                                vjust = 1,
                                                hjust=1),
                     plot.title = element_text(size=11),
                     legend.text = element_text( size=8),
                     legend.title = element_text( size=9))

```


```{r}
############################################################
###############       DATA LOAD             ############### 
############################################################

imdb <- read.csv("IMDB.csv", stringsAsFactors = F, sep=";")


data.frame(variable = names(imdb),
           class = sapply(imdb, class),
           first_values = sapply(imdb, function(x) paste0(head(x),collapse = ", ")),
           row.names = NULL)
```

We first check for any missing values and see that there are no NAs.

```{r}
summary(imdb)
```

Given the range of gross and budget we can switch to working in unit numbers by dividing by a million.

```{r}
imdb<- imdb%>% 
        mutate(gross = gross/1000000,
               budget = budget/1000000)
```

# Exploratory Data Analysis

We are interested in predicting the gross of a movie basic on its 
characteristics. First let's analyze the target variable.

```{r}
basicStats(imdb%>%dplyr::select(gross))
```

Using the basicStats we obtain the excess kurtosis, $K(X)-3$ and we see that we 
have a considerable positive one and that it has a right skewed distribution.
So, it is not normal. We should consider that the skewness and kurtosis could be
due to outliers.

We look at an overview of the relationship between all variables in our dataset:

```{r}
pairs(~.,imdb %>% select(-c(movietitle,genre, titleyear)))
```
In this plot we observe that some variables seem to be correlated, such as $actor1fl$ with $castfl$, as well as, $budget$ with $duration$. However, this correlation would present a problem, in the form of multicolinearity, in the 
case that both variables were to be included in the final model.

We now look closer into the relation between $gross$ and all the numerical 
variables.
```{r}
imdb %>% 
  select(-c(movietitle,genre, titleyear)) %>%
  gather(-gross, key = "some_var_name", value = "some_value_name") %>%
  ggplot(aes(x = some_value_name, y = gross)) +
  geom_point() +
  facet_wrap(~ some_var_name, scales = "free")+
  ggstyleFonts +
  theme(panel.grid.major = element_blank(),
        axis.title.x = element_blank())
```

We observe more of a linear relation between the pairs of gross and budget, as 
well as with, duration. We can't discern any pattern between the pairs of gross 
and the Facebook variables: $directorfl,actor1fl,actor2fl,actor3fl,castfl$. 
The $actor3fl,directorfl$ could be separated in 2 clusters at the cutoff point 
of 5000 likes and for the latter at the cutoff point of 10000 likes.

We create a categorial variable ($yearcat$) with 3 levels: 2000-2005, 2006-2010 
and 2011-2016 based on the $titleyear$ of the movie.

```{r}
imdb <- imdb%>%
        mutate(yearcat = as.factor(
                          ifelse(titleyear< 2006, "2000-2005", 
                          ifelse(titleyear< 2011, "2006-2010","2011-2016" ))
                          ),
               genre = as.factor(genre))
ggplot(imdb%>%
         group_by(titleyear,yearcat)%>% 
         summarise(movies = n()) ,
       aes(x=titleyear, y=movies, fill= yearcat))+
  geom_bar(stat="identity") +
  geom_text(aes(label= movies), 
            position=position_stack(vjust=0.5), colour="white" ,size=3) +
  scale_fill_d3(name="") +
  labs(y = "# movies", x= "year" , title = "Cluster movies into 3 categories by year") +
  ggstyle+
  theme(legend.position="bottom")
```


```{r}
imdb%>%
  group_by(titleyear,yearcat)%>%
  summarise(movies = n()) %>%
  group_by(yearcat) %>%
  mutate(avgMovies = mean(movies)) %>%
  summarise(movies = sum(movies),
            avgMovies = max(avgMovies)) %>%
  mutate(pcn = movies/sum(movies))
```

The movies are roughly uniformly distributed between the three categories.
However, on average more movies were released between the years 2006 and 2010. 
In addition, based on the significant difference between 2016 and all the previous
years it is highly probable that we don't have data for the whole year. 
So, we have two categorical variables: the year category and the genre. 
Let's see how the economical variables relates to $genre$. 

```{r,message = FALSE}
imdb %>% 
  select(c(budget,duration,gross,genre)) %>% 
  ggpairs(.,
          title = "Imdb economical variables relation by genre", 
          mapping = ggplot2::aes(colour=genre), 
          lower = list(#continuous = wrap("smooth", alpha = 0.3, size=0.1),
                      continuous = wrap("points", size=0.1),
                      discrete = "blank", combo="blank"), 
          diag = list(discrete="barDiag", 
                      continuous = wrap("densityDiag", alpha=0.5 )), 
          upper = list(combo = wrap("box_no_facet", alpha=0.5),
                      continuous = wrap("cor", size=2, alignPercent=0.8))) +
  ggstyle+
  theme(panel.grid.major = element_blank())
```

We observe two outliers in the Action genre based on their $gross$ value, which turn out to be blockbusters.
```{r}
imdb %>% dplyr::filter(gross> 600) %>% pull(movietitle)
```

The distribution of $gross$ for the Action genre is skewed to the right and has 
a higher IQR than the rest of the genres. However, it is also the genre with 
the smallest number of movies.
Similarly, $budget$ has excess kurtosis with more heavier tails than $gross$
especially for the action movies.
In the linear relation that we observed before between $gross$ and $budget$ we 
add now the genre which confirms this relation, particularly more for the Action movies.

```{r,message = FALSE}
imdb %>% 
  select(c(budget,duration,gross,yearcat)) %>% 
  ggpairs(.,
          title = "Imdb economical variables relation by Year group", 
          mapping = ggplot2::aes(colour=yearcat), 
          lower = list(#continuous = wrap("smooth", alpha = 0.3, size=0.1),
            continuous = wrap("points", size=0.1),
            discrete = "blank", combo="blank"), 
          diag = list(discrete="barDiag", 
                      continuous = wrap("densityDiag", alpha=0.5 )), 
          upper = list(combo = wrap("box_no_facet", alpha=0.5),
                       continuous = wrap("cor", size=2, alignPercent=0.8))) +
  ggstyle+
  theme(panel.grid.major = element_blank())
```

On the other hand, we don't observe any differences between the different years.

# Fit complete model

We first fit the complete model including as predictors, all the numerical
variables, the two categorical variables, the categorical-categorical
interactions and the interaction between numerical-categorical. 

```{r}
rownames(imdb) <- imdb$movietitle
imdb <- imdb %>% dplyr::select(-c(movietitle,titleyear))
mc<-lm(gross~.*(genre+yearcat), imdb)

glance(mc)
```

Roughly 64% of the variance found in the response variable (*gross*) can be
explained by the predictor variables. The obtained p-value (*Omnibus test*)
indicates that the overall model is significant. 

```{r, warning=FALSE}
op<-par(mfrow=c(2,2))
plot(mc)
par(op)
```

From the *Normal Q-Q* plot we see that there is assymetry in the distribution
and we can conclude that normality of the residuals is not met. From the 
*Scale vs Location* plot, we seek to validate the assumption of 
homoskedasticity, which does not seem to hold in our case. What's more, from the
*Residual vs Fitted* plot we observe, a non random distribution of the points
along the *y-axis*. All in all, we can't validate this model. We look into this
with more detail with the final model.

# Select significant variables

We use the stepwise procedure, by using the $BIC$ criterion, to select the
significant variables. Since our objective is the interpretability of the model we
choose as starting point the null model, in contrast to starting 
form the complete.

```{r}
m0 <- lm(gross~1, imdb)
summary(m1<-step(m0,scope=list(upper=mc), direction="both",
                 k=log(nrow(imdb)), trace = 0))
```

Contrary to the complete model, we see that just 60% of the variance can be
explained, although the obtained p-value indicates that the overall model is
significant. However, we see that neither *actor1fl*, *actor2fl*, *castfl*,
*directorfl*, *facenumber_in_poster* nor *yearcat* are included. For this
reason, we consider exploring the stepwise procedure starting from the complete
model.

```{r}
summary(m1<-step(mc,direction="both",k=log(nrow(imdb)), trace = 0))
```

Similarly to the complete model, we see that roughly 63% of the variance can be
explained and the obtained p-value indicates that the overall model is
significant. 
In this case, we see that *actor3fl* is not included but *actor1fl*, *actor2fl*,
*castfl* and *yearcat* are . Likewise, *directorfl* and *facenumber_in_poster*
are not included in the model. We decide to continue our analysis with this 
last model.

When dealing with categorical variables we should use the $Anova$ method. The
$p-value$ obtained will allow us to say if the interaction variables are
significant. 

```{r}
car::Anova(m1)
```

We see that the interaction variables *budget:yearcat*, *duration:genre*,
*actor1fl:genre* and *castfl:genre* are significant, so we keep them in our
model.
Although the variable *yearcat* seems to not be significant, we decide to keep
it in our model due to its interaction being significant.

# Check for multicollinearity

Strong associations between predictors will increase standard errors, and
therefore increase the probability of a type-II error, as well as affect the
value of the coefficients. In order to detect it in our model, the diagnostic
that we will use is the variance-inflation factor.

```{r}
car::vif(m1)
```

From the normalized inflation factor (second column in the previous output) we
conclude that *actor1fl* and *castfl* may be causing multicollinearity in the
model. This could be perhaps due to a correlation between them as we saw in the
exploratory analysis. Consequently, the coefficients can't be directly
interpreted.

By definition a categorical variable that is included in an interaction term (as
well as the interaction terms themselves) will have a high VIF factor and hence
there is no reason to further investigate.

The following figure confirms our suspicions about the correlated variables:

```{r}
corr_mat=cor(imdb %>% 
              select_if(is.numeric) %>%
               select(-c(gross)),method="s")
corrplot::corrplot(corr_mat,type = "lower", method = "number", 
                  tl.col = "black", tl.cex = 0.5,number.cex = 0.7)
```

To proceed we would need to select which of this three correlated variables
(*actor1fl*, *actor2fl* or *castfl*) would result in a better model and whether
the VIF factor will be corrected.

```{r}
lm.actor1 <- update(m1,.~.-(castfl+actor2fl+castfl:genre))
lm.actor2 <- update(m1,.~.-(castfl+actor1fl+castfl:genre+actor1fl:genre))
lm.cast <- update(m1,.~.-(actor1fl+actor2fl+actor1fl:genre))

glance(lm.actor1)
car::vif(lm.actor1)
glance(lm.actor2)
car::vif(lm.actor2)
glance(lm.cast)
car::vif(lm.cast)
```

Comparing the $R^2$ between the 3 models we see that we do not obtain a
siginificantly better model with any of the variables. On the other hand,
looking at the change in the VIF we can conclude that best correction is
obtained by the *castfl* model. Given that we lack expert domain knowledge to
guide us we decide to keep *castfl* as it is an added variable of the likes of
the whole movie cast (it includes the other measures).

```{r}
m1 <- update(m1,.~.-(actor1fl+actor2fl+actor1fl:genre))
```

# Validate model's assumptions

```{r, warning=FALSE}
op<-par(mfrow=c(2,2))
plot(m1)
par(op)
```

From the *Normal Q-Q* plot we see that there is still assymetry in the
distribution and we can conclude that normality of the residuals is not met.

To look at the *Residual vs Fitted* plot in more detail, we plot it for each
predictor.

```{r}
car::residualPlots(m1)
```

Looking at both the figures, as well as, the curvature test, we conclude that a
quadratic term is not needed for any of the variables.

From the *Scale-Location* plot, we seek to validate the assumption of
homoskedasticity, which does not seem to hold in our case.
Consequently, we consider a log transformation to both *gross* and *budget*
measures and see whether the obtained model better meets the assumptions.

```{r}
mlog <- lm(log10(gross) ~ log10(budget) + duration + castfl + genre + yearcat +
             budget:yearcat + duration:genre + castfl:genre,
           data = imdb)
glance(mlog)
op<-par(mfrow=c(2,2))
plot(mlog)
par(op)
```

We now have better agreement with the model's assumptions at the cost of a
decrease in the model's $R^2$.

Regarding the *Residuals vs Leverage* plot, we see that with our regular final
model, without the logs, there are some points that are influential and could be
considered to be outliers. However, in the transformed model they do not appear
to be so.

Having mentioned that, we decide to keep the previous model as it has a better
fit with our data and a more intuitive interpretation.

# Model interpretation

For the model interpretation, we will use the *effect* plots to interpret the coefficients of our final model, taking into account the interaction terms.

In first place, we look at the effect of the budget variable.
```{r}
m1.effects <- effects::allEffects(m1)
plot(m1.effects, "budget:yearcat")
```

We clearly see a similar linear increase across all categories of the *yearcat* variable, being almost identical for *2000-2005* and *2006-2010* and slightly flatter for *2011-2016*. This means that the more budget a movie has, the more probable it is it has a bigger revenue. Specifically, it means that for every additional million dollars of budget we have an increase of `r m1$coefficients["budget"]`, `r m1$coefficients["budget"] + m1$coefficients["budget:yearcat2006-2010"]` and `r m1$coefficients["budget"] + m1$coefficients["budget:yearcat2011-2016"]` million dollars in gross revenue, respectively for each category.

```{r}
plot(m1.effects, "duration:genre")
```

Overall, the duration of Drama and Terror movies cause a lesser increase to the gross revenue than Action and Comedy. Specifically, for every increase in 1 minute the gross revenue increases `r m1$coefficients["duration"] + m1$coefficients["duration:genreDrama"]` and `r m1$coefficients["duration"] + m1$coefficients["duration:genreTerror"]` million dollars for Drama and Terror respectively. On the other hand, for every increase in 1 minute in Action and Comedy movies gross revenue increases `r m1$coefficients["duration"]` and `r m1$coefficients["duration"] + m1$coefficients["duration:genreComedy"]` million dollars respectively.

```{r}
plot(m1.effects, "castfl:genre")
```

Contrary to the effect of *duration*, we now sauses a higher increase to the gross revenue compared to Terror and Comedy. Specifically, for every thousand cast likes the gross revenue increases `r 1000*m1$coefficients["castfl"]` and `r 1000*m1$coefficients["castfl"] + 1000*m1$coefficients["castfl:genreDrama"]` million dollars for Action and Drama respectively. On the other hand, for every thousand cast likes in Comedy and Terror movies gross revenue increases `r 1000*m1$coefficients["castfl"] + 1000*m1$coefficients["castfl:genreComedy"]` and `r 1000*m1$coefficients["castfl"] + 1000*m1$coefficients["castfl:genreTerror"]` million dollars respectively.

In conclusion, the defining attributes that make up a movie (i.e. budget, duration and cast) seem to have a time invariant, but genre dependent, relation to the gross revenue. 
