---
title: | 
  | \LARGE ASM Practice 
subtitle: "GAMs for hirsutism data"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "06/01/2020"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
set.seed(42)
```

```{r, results="hide"}
library(sm)
library(rgl)
library(mgcv)
library(dplyr)

hirs <- read.table("hirsutism.dat",header=T, sep="\t",fill=TRUE)
hirs <- hirs %>% tidyr::drop_na()
hirs$Treatment <- as.factor(hirs$Treatment)
attach(hirs)
```

A clinical trial was conducted to evaluate the effectiveness of an antiandrogen 
combined with an oral contraceptive in reducing hirsutism for 12 consecutive months.
The data set `hirsutism.dat` contains artificial values of measures corresponding to 
some patients in this study. The variables are the following:

* `Treatment`, with values 0, 1, 2 or 3.
* `FGm0`, it indicates the baseline hirsutism level at the randomization moment 
(the beginning of the clinical trial). Only women with baseline FG values grater 
than 15 where recruited.
* `FGm3`, FG value at 3 months.
* `FGm6`, FG value at 6 months.
* `FGm12`, FG value at 12 months, the end of the trial. 
* `SysPres`, baseline systolic blood pressure.
* `DiaPres`, baseline diastolic blood pressure.
* `weight`, baseline weight.
* `height`, baseline height.

Our objective is to fit several *GAM* models (including semiparametric models) 
explaining `FGm12` as a function of the variables that were measured at the 
beginning of the clinical trial (including `FGm0`) and `Treatment` (treated as
factor).

Before proceeding to the model fitting, we remove the rows which have missing
values, which are 8 rows.

During the following discussion, we use the function `gam` from package `mgcv` to fit our additive models.

# Baseline linear model

After fitting a complete linear model and removing the non-significant terms, we get a model with the followig formula:

* `FGm12 ~ FGm0 + Treatment`

```{r}
# Linear Model
am.lin <- gam(FGm12 ~ FGm0 + Treatment,
             data=hirs)
summary(am.lin)
```

Which we use as a baseline model to compare all following additive models.

# Linear vs smooth terms

We want to explore whether our predictors should enter the model as linear
parametric or smooth terms.
To see this, we fit a model with all numerical predictors as smooth terms, 
having the following formula:

* `FGm12 ~ Treatment + s(FGm0) + s(SysPres) + s(DiaPres) + s(weight) + s(height)`

```{r}
# Additive model 1
am.1 <- gam(FGm12 ~ Treatment + s(FGm0) + s(SysPres) + s(DiaPres) 
                + s(weight) + s(height),
                data=hirs)
summary(am.1)
plot(am.1,pages=1,residuals=TRUE)
```

From the summary output we see that the effective degrees of freedom (edf) of 
the *SysPres*, *weight* and *height* predictors are close to 1. In addition, 
from the plots we see that the form of the effect of these variables is linear.
Therefore, we include these terms linearly and refit.

```{r}
# Additive model 2
am.2 <- gam(FGm12 ~ s(FGm0) + Treatment + SysPres + s(DiaPres) + weight + height,
                data=hirs)
summary(am.2)
plot(am.2,pages=1,residuals=TRUE)
```

As expected the model does not change when we turn the smooth terms that had edf close to 1 into linear terms.

We know compare this model with the previous linear baseline.

```{r}
anova(am.lin, am.2, test="F")
```

Although the models do not seem to differ greatly, the explained deviance of 
the smooth term model, as well as the adjusted $R^2$ value, is higher.
Therefore, we prefer the model with smooth terms.

We see a visualization of the predicted values depending on the predictors *Treatment* and *FGm0*:

```{r}
vis.gam(am.2, view = c("Treatment", "FGm0"), theta=60, phi=20, zlab = "FGm12")
```

We see for the Treatments 1 to 3 the spline of *FGm12* depending on *FGm12* is
almost the same, while for treatment 0 (only contraceptives) all values of *FGm12*
are higher independent on the initial values of *FGm0*.

# Smoothing terms by Treatment

In order to better describe the behaviour of the predictors depending on the *Treatment* value, we can specify the fitting on different smooth terms for each category by using the formula:

* `FGm12 ~ s(FGm0, by= Treatment) + Treatment + SysPres + s(DiaPres) + weight + height`

```{r}
# Additive model 3
am.3 <- gam(FGm12 ~ s(FGm0, by= Treatment) + Treatment + SysPres + s(DiaPres) 
            + weight + height,
            data=hirs)
summary(am.3)
plot(am.3,pages=1,residuals=TRUE)
```

```{r}
anova(am.2, am.3, test="F")
```

```{r}
vis.gam(am.3, view = c("Treatment", "FGm0"), theta=30, phi=30, zlab = "FGm12")
```

```{r}
# Additive model 4
am.4 <- gam(FGm12 ~ s(FGm0, by= Treatment) + Treatment, data=hirs)
summary(am.4)
plot(am.4,pages=1,residuals=TRUE)
```

```{r}
anova(am.3, am.4, test="F")
```

```{r}
vis.gam(am.4, view = c("Treatment", "FGm0"), theta=30, phi=30, zlab = "FGm12")
```






# Additive Models

We use function `gam` from package `mgcv` to fit the following additive models:

* `FGm12 ~ FGm0 + Treatment`
* `FGm12 ~ s(FGm0) + Treatment`
* `FGm12 ~ s(FGm0, by= Treatment)`


```{r}
# Linear Model
am1.0 <- gam(FGm12 ~ FGm0 + Treatment, data=hirs)
am1.0
summary(am1.0)
```

```{r}
# Additive model 1
am1.1 <-  gam(FGm12 ~ s(FGm0) + Treatment, data=hirs)
summary(am1.1)
plot(am1.1,pages=1,residuals=TRUE)
```

```{r}
# Additive model 2
am1.2 <-  gam(FGm12 ~ s(FGm0, by=Treatment), data=hirs)
summary(am1.2)
plot(am1.2,pages=1,residuals=TRUE)
```

## ANOVA type tests

We use function `anova` to compare the fitted models two by two.

```{r}
anova(am1.0,am1.1,test="F")
```

```{r}
anova(am1.0,am1.2,test="F")
```

```{r}
anova(am1.1,am1.2,test="F")
```

## Explaining `FGm12` by SysPres and DiaPres

* `FGm12 ~ s(FGm0) + Treatment + SysPres + DiaPres`
* `FGm12 ~ s(FGm0) + Treatment + s(SysPres) + DiaPres`
* `FGm12 ~ s(FGm0) + Treatment + SysPres + s(DiaPres)`
* `FGm12 ~ s(FGm0) + Treatment + s(SysPres) + s(DiaPres)`
* `FGm12 ~ s(FGm0) + Treatment + s(SysPres, DiaPres)`


```{r}
# Additive model 3
am1.3 <-  gam(FGm12 ~ s(FGm0) + Treatment + SysPres + DiaPres, data=hirs)
summary(am1.3)
plot(am1.3,pages=1,residuals=TRUE)
```

```{r}
# Additive model 4
am1.4 <-  gam(FGm12 ~ s(FGm0) + Treatment + s(SysPres) + DiaPres, data=hirs)
summary(am1.4)
plot(am1.4,pages=1,residuals=TRUE)
```

```{r}
# Additive model 3
am1.5 <-  gam(FGm12 ~ s(FGm0) + Treatment + SysPres + s(DiaPres), data=hirs)
summary(am1.5)
plot(am1.5,pages=1,residuals=TRUE)
```

```{r}
# Additive model 6
am1.6 <-  gam(FGm12 ~ s(FGm0) + Treatment + s(SysPres) + s(DiaPres), data=hirs)
summary(am1.6)
plot(am1.6,pages=1,residuals=TRUE)
```


```{r}
# Additive model 3
am1.7 <-  gam(FGm12 ~ s(FGm0) + Treatment + s(SysPres, DiaPres), data=hirs)
summary(am1.7)
plot(am1.7,pages=1,residuals=TRUE)
```

### ANOVA type tests

We use function `anova` to compare the fitted models two by two.

```{r}
anova(am1.3,am1.4,test="F")
```

```{r}
anova(am1.3,am1.5,test="F")
```

```{r}
anova(am1.3,am1.6,test="F")
```

```{r}
anova(am1.3,am1.7,test="F")
```

```{r}
anova(am1.4,am1.5,test="F")
```

```{r}
anova(am1.6,am1.7,test="F")
```


## Explaining `FGm12` by `weight` and `height`

* `FGm12 ~ s(FGm0) + Treatment + weight + height`
* `FGm12 ~ s(FGm0) + Treatment + s(weight) + height`
* `FGm12 ~ s(FGm0) + Treatment + weight + s(height)`
* `FGm12 ~ s(FGm0) + Treatment + s(weight) + s(height)`
* `FGm12 ~ s(FGm0) + Treatment + s(weight, height)`

```{r}
# Additive model 3
am2.3 <-  gam(FGm12 ~ s(FGm0) + Treatment + weight + height, data=hirs)
summary(am2.3)
plot(am2.3,pages=1,residuals=TRUE)
```

```{r}
# Additive model 4
am2.4 <-  gam(FGm12 ~ s(FGm0) + Treatment + s(weight) + height, data=hirs)
summary(am2.4)
plot(am2.4,pages=1,residuals=TRUE)
```

```{r}
# Additive model 5
am2.5 <-  gam(FGm12 ~ s(FGm0) + Treatment + weight + s(height), data=hirs)
summary(am2.5)
plot(am2.5,pages=1,residuals=TRUE)
```


```{r}
# Additive model 6
am2.6 <-  gam(FGm12 ~ s(FGm0) + Treatment + s(weight) + s(height), data=hirs)
summary(am2.6)
plot(am2.6,pages=1,residuals=TRUE)
```


```{r}
# Additive model 7
am2.7 <-  gam(FGm12 ~ s(FGm0) + Treatment + s(weight, height), data=hirs)
summary(am2.7)
plot(am2.7,pages=1,residuals=TRUE)
```

### ANOVA type tests

We use function `anova` to compare the fitted models two by two.

```{r}
anova(am2.3,am2.4,test="F")
```

```{r}
anova(am2.3,am2.5,test="F")
```

```{r}
anova(am2.3,am2.6,test="F")
```

```{r}
anova(am2.3,am2.7,test="F")
```

```{r}
anova(am2.4,am2.5,test="F")
```

```{r}
anova(am2.6,am2.7,test="F")
```
