---
title: | 
  | \LARGE ASM Homework 2
subtitle: "Generalized Linear Model for JYB data"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "22/10/2019"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE,results="hide"}

############################################################
###############  libraries  & Functions      ############### 
############################################################

### load packages ###
requireorinstall=function(package=""){
  reqpac=parse(text=paste("require(",as.character(package),")"))
  if(eval(reqpac)){
    print(paste(as.character(package), "has been loaded correctly"))
  } else {
    print(paste("trying to install" ,as.character(package)))
    eval(parse(text=paste("try(install.packages(",as.character(package),"))")))
    if(eval(reqpac)){
      print(paste(as.character(package) ,"has been installed and loaded correctly"))
    } else {
      warning(paste("could not install",as.character(package)))
    }
  }
}


requireorinstall(c("knitr","kableExtra","ggplot2","ggsci","dplyr","tidyr","lmtest",
                   "tidyverse","broom","purrr","magrittr","grid","gridExtra","scales",
                   "GGally","fBasics", "car", "effects", "emmeans", "viridis"))

## theme definition for ggplot 
ggstyle = theme(panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_blank(),
                axis.line = element_blank(),
                axis.title.x = element_text(size = 10,
                                            vjust=0.1),
                axis.title.y = element_text(size = 10,
                                            vjust=1.5),
                axis.text.y = element_text(size=9),
                axis.text.x = element_text(size=9,
                                           angle = 45,
                                           vjust = 1,
                                           hjust=1),
                plot.title = element_text(size=11),
                legend.text = element_text( size=8),
                legend.title = element_text( size=9))

ggstyleFonts = theme(axis.title.x = element_text(size = 10,
                                                 vjust=0.1),
                     axis.title.y = element_text(size = 10,
                                                 vjust=1.5),
                     axis.text.y = element_text(size=9),
                     axis.text.x = element_text(size=9,
                                                angle = 45,
                                                vjust = 1,
                                                hjust=1),
                     plot.title = element_text(size=11),
                     legend.text = element_text( size=8),
                     legend.title = element_text( size=9))

```

# Exploratory Data Analysis

We check for any missing values or attributes without a value and find none nor NAs.

```{r,warning=FALSE,echo = FALSE,results="asis"}
jyb <- read.csv("JYB.csv", sep=";")

dfNum <- jyb %>% dplyr::select_if(is.numeric) %>% dplyr::select(-id)

t1<-data.frame(variable = names(dfNum),
           class = sapply(dfNum, class),
           min = sapply(dfNum, function(x) min(x)),
           mean = sapply(dfNum, function(x) mean(x)),
           median =  sapply(dfNum, function(x) median(x)),
           max = sapply(dfNum, function(x) max(x)),
           row.names = NULL)   %>%
  kable(format = "latex", booktabs = TRUE,digits = 2) %>% 
  kable_styling(position = "center",font_size = 9,
                latex_options = c("HOLD_position"))

t2<-jyb %>% 
  dplyr::select_if(is.factor) %>% 
  map(levels) %>% 
  map(length) %>% 
  unlist(recursive = FALSE) %>% 
  enframe() %>% 
  kable(format = "latex", booktabs = TRUE,
       col.names =c("attribute", "# levels"))%>%
  kable_styling(position = "center",font_size = 9,
                latex_options = c("HOLD_position"))

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.5\\linewidth}
      \\caption{Numerical variables}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.5\\linewidth}
      \\centering
        \\caption{Categorical variables}",
        t2,
    "\\end{minipage} 
\\end{table}"
))
```


```{r,warning=FALSE,echo = FALSE,results="asis"}
plyr::ldply(jyb %>% dplyr::select_if(is.factor) %>%
              map(levels) , rbind) %>%
  mutate_all(funs(fct_explicit_na(., na_level = ""))) %>%
  kable(col.names= c("attribute",paste0(rep("level_",4),1:12)),
        format = "latex", booktabs = TRUE)%>%
  kable_styling(latex_options = "scale_down")
```

We are interested in predicting whether the customer subscribed to the deposit,
so our target variable *y* is a binary one. 

We now look closer into the relation between *y* and all the numerical 
variables.
```{r, echo = FALSE,fig.align="center",out.width = "300pt"}
jyb %>% 
  dplyr::select(c("y",colnames(jyb %>% 
                                 dplyr::select_if(is.numeric) %>% 
                                 dplyr::select(-id)))) %>%
  gather(-y, key = "some_var_name", value = "some_value_name") %>%
  ggplot(aes(x = some_value_name, fill = y)) +
  geom_density( alpha = .2) +
  scale_fill_viridis(name="y",option="plasma",  discrete = TRUE) +
  facet_wrap(~ some_var_name, scales = "free")+
  ggstyleFonts +
  theme(panel.grid.major = element_blank(),
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = '#ffffff'),
        legend.position="bottom")

```

# Complete Model


# Evaluation First Order Interactions


# Automatic Variable Selection process


# Model comparison


# Model validation


# Model interpretation