---
title: | 
  | \LARGE ASM Practice 
subtitle: "Smoothing and regression splines"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "14/12/2019"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
set.seed(42)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, results="hide"}
requireorinstall=function(package=""){
  reqpac=parse(text=paste("require(",as.character(package),")"))
  if(eval(reqpac)){
    print(paste(as.character(package), "has been loaded correctly"))
  } else {
    print(paste("trying to install" ,as.character(package)))
    eval(parse(text=paste("try(install.packages(",as.character(package),"))")))
    if(eval(reqpac)){
      print(paste(as.character(package) ,"has been installed and loaded correctly"))
    } else {
      warning(paste("could not install",as.character(package)))
    }
  }
}
requireorinstall(c("splines","tidyverse","dplyr","kable","kableExtra"))
```

# 1
Estimate the regression function m(instant) of cnt as a function of instant
using a cubic regression splines estimated with the R function smooth.splines
and choosing the smoothing parameter by Generalized Cross Validation.

```{r, echo=FALSE, warning=FALSE}
bikes <-  get(load("bikes.Washington.Rdata"))
#  "generalized" cross-validation (GCV) when cv =FALSE;
sm.sp.1 <- smooth.spline(x = bikes$instant, y = bikes$cnt,
                         cv = FALSE, all.knots = FALSE)
```

## a) 
The chosen Smoothness penalization hyperparameter $\lambda$ by GCV is 
`r sm.sp.1$lambda`.

## b) 
The corresponding equivalent number of degrees of freedom of the spline
regression's linear estimator is `r round(sm.sp.1$df)`.

## c) 
`r length(sm.sp.1$fit$knot)` knots were used.

## d) 
We show a scatterplot of the data points with the fitted spline regression:

```{r, out.width='350pt', fig.align='center'}
plot(bikes$instant,bikes$cnt, col="black", ylab = "count", xlab = "instant")
lines(sm.sp.1, lty = 1, lw=2, col = "blue")
```

## e) 
We estimate now m(instant) by unpenalized regression splines combining
the R functions *bs* and *lm*, using the knots where *n.knots* is the previous 
value of df minus 4.
```{r, echo=TRUE}
x <- bikes$instant # x
y <- bikes$cnt # y
n <- length(x)
n.knots <- sm.sp.1$df -4
my.knots <- quantile(x,((1:n.knots)-.5)/n.knots) 
b.kn <- range(x)+c(-1,1)*.1*(diff(range(x)))
X.bs <- bs(x, knots=my.knots, Boundary.knots = b.kn)
sreg <- lm(y~X.bs)
```

## f) 
Plot the scatter plot with the different spline regressions

```{r, out.width='350pt', fig.align='center'}
{
  plot(bikes$instant,bikes$cnt, col="black", xlab = "instant", ylab = "count")
  lines(sm.sp.1, lty = 1, lw=2, col = "blue")
  lines(bikes$instant, sreg$fitted.values, lty = 1, lw=2, col = "red")
  legend("topleft", col=c("blue","red"), lw=2,
         legend = c("Smooth Spline Regression","Unpenalized Spline Regression"))
}
```

# 2


We define a new variable cnt.5000 taking the value 1 for days such that the number 
of total rental bikes is larger than or equal to 5000, or 0 otherwise.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results="asis"}
source('IRWLS_logistic_regression.R')

bikes <- bikes %>% mutate(cnt.5000 = ifelse(cnt>=5000, 1,0))

bikes%>% group_by(cnt.5000) %>% summarise(n= n()) %>%
  kable(format = "latex", booktabs = TRUE,digits = 2) %>% 
  kable_styling(position = "center",font_size = 9,
                latex_options = c("HOLD_position"))
```


## a)

We use the function *logistic.IRWLS.splines* to fit the non-parametric binary 
regression *cnt.5000* as a function of the temperature, using *df=6*. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}

my.spline.glm  <- logistic.IRWLS.splines(x = bikes$temp, 
                                         y = bikes$cnt.5000, df= 6,
                                          all.knots=TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.align='center', out.height="300pt"}
plot(bikes$temp,bikes$cnt.5000,col=2,xlab="temp",ylab="cnt5000",
     main="Non-parametric binary regression cnt.5000\n(with fitted values) as a function of the temperature")
lines(bikes$temp,my.spline.glm$fitted.values,col="purple")
abline(h = 0.5, lty =1)
abline(v = min(bikes$temp[my.spline.glm$fitted.values>0.5]), lty =2)
abline(v = max(bikes$temp[my.spline.glm$fitted.values>0.5]), lty =2)
```

The range of temperatures that the $Pr(cnt >= 5000|temp)$ is larger than $0.5$ 
is from `r round(min(bikes$temp[my.spline.glm$fitted.values>0.5]),1)`$^\circ C$ to 
`r round(max(bikes$temp[my.spline.glm$fitted.values>0.5]),1)`$^\circ C$.


## b) 

We now choose the parameter df by *k-fold* cross validation with $k = 5$ and using 
$df.v = 3:15$ as the set of possible values for *df*.

```{r}
df.v = 3:15
cv = 5
out.cv <- data.frame(df.v=df.v, CV_error=rep(0, length(df.v)))
folds <- caret::createFolds(bikes$temp, k=cv)
for (i in 1:length(df.v)) {
    error <- rep(0, cv)
    for(j in 1:cv) {
        fold <- folds[[j]]
        x.train <-bikes$temp[-fold]
        y.train <-bikes$cnt.5000[-fold]
        x.test <-bikes$temp[fold]
        y.test <-bikes$cnt.5000[fold]
        my.spline.glm  <- logistic.IRWLS.splines(x = x.train, 
                                                 y = y.train,
                                                 x.new = x.test,
                                                 df= df.v[i],
                                                 all.knots=TRUE)
        pred <-ifelse(my.spline.glm$predicted.values>=0.5, 1,0)
        confMx <- caret::confusionMatrix(data = as.factor(pred),
                                         reference= as.factor(y.test))
        error[j] <- 1-sum(diag(confMx$table))/length(y.test)
    }
    out.cv$CV_error[i] <- mean(error)
}
```

```{r, fig.align='center', out.height="300pt"}
plot(out.cv$df.v,out.cv$CV_error, type = "b", 
     xlab = "degrees of freedom", ylab = "CV error rate")

```

The minimum is obtained at `r out.cv$df.v[which.min(out.cv$CV_error)]` degrees of freedom. 
We now refit the non-parametric binary regression cnt.5000 as a function of the temperature
using the obtained df. 

```{r}

dfmin <- out.cv$df.v[which.min(out.cv$CV_error)]

my.spline.glm  <- logistic.IRWLS.splines(x = bikes$temp, 
                                         y =bikes$cnt.5000,
                                         df= dfmin,
                                         all.knots=TRUE)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.align='center', out.height="300pt"}
plot(bikes$temp,bikes$cnt.5000,col=2,xlab="temp",ylab="cnt5000",
     main="Non-parametric binary regression cnt.5000\n(with fitted values) as a function of the temperature")
lines(bikes$temp,my.spline.glm$fitted.values,col="purple")
abline(h = 0.5, lty =1)
abline(v = min(bikes$temp[my.spline.glm$fitted.values>0.5]), lty =2)
abline(v = max(bikes$temp[my.spline.glm$fitted.values>0.5]), lty =2)
```

The range of temperatures that the $Pr(cnt >= 5000|temp)$ is larger than $0.5$ 
is from `r round(min(bikes$temp[my.spline.glm$fitted.values>0.5]),1)`$^\circ C$ to 
`r round(max(bikes$temp[my.spline.glm$fitted.values>0.5]),1)`$^\circ C$.