---
title: | 
  | \LARGE ASM Practice 
subtitle: "Ridge Regression"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "27/10/2019"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, results="hide"}
requireorinstall=function(package=""){
  reqpac=parse(text=paste("require(",as.character(package),")"))
  if(eval(reqpac)){
    print(paste(as.character(package), "has been loaded correctly"))
  } else {
    print(paste("trying to install" ,as.character(package)))
    eval(parse(text=paste("try(install.packages(",as.character(package),"))")))
    if(eval(reqpac)){
      print(paste(as.character(package) ,"has been installed and loaded correctly"))
    } else {
      warning(paste("could not install",as.character(package)))
    }
  }
}
requireorinstall(c("GGally","broom","car","caret","dplyr","ggplot2","ggsci","grid","gridExtra","kableExtra","knitr","lmtest","purrr","scales","tidyr","tidyverse"))
## theme definition for ggplot 
ggstyle = theme(panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_blank(),
                axis.line = element_blank(),
                axis.title.x = element_text(size = 10,
                                            vjust=0.1),
                axis.title.y = element_text(size = 10,
                                            vjust=1.5),
                axis.text.y = element_text(size=9),
                axis.text.x = element_text(size=9,
                                           angle = 45,
                                           vjust = 1,
                                           hjust=1),
                plot.title = element_text(size=11),
                legend.text = element_text( size=8),
                legend.title = element_text( size=9))
ggstyleFonts = theme(axis.title.x = element_text(size = 10,
                                                 vjust=0.1),
                     axis.title.y = element_text(size = 10,
                                                 vjust=1.5),
                     axis.text.y = element_text(size=9),
                     axis.text.x = element_text(size=9,
                                                angle = 45,
                                                vjust = 1,
                                                hjust=1),
                     plot.title = element_text(size=11),
                     legend.text = element_text( size=8),
                     legend.title = element_text( size=9))

```

# Choosing the penalization parameter $\lambda$


# Ridge regression for the Boston Housing data

We start by scaling and splitting the Boston dataset to training and test using a 2/3 ratio. 
Since *CHAR* is a factor variable we do not include it in the *scale* function.
First we need to tune the parameter $\lambda$. To do this we use 10 fold corss validation
performed by *cv.glmnet*. 

```{r, echo = FALSE, message=FALSE, warning=FALSE,fig.align="center",out.width = "300pt"}
############################################################
###############       DATA LOAD             ############### 
############################################################

load("boston.Rdata")
# not including TOWNNO,TOWN, LON, LAT, CMEDV
# since statement asks to fit the regression model where the response is MEDV 
# and the explanatory variables are the remaining 13 variables in the previous list

boston <- boston.c %>% 
            dplyr::select(CHAS,CRIM, ZN,INDUS, NOX, RM,AGE,DIS, RAD, TAX,
                          PTRATIO,B,LSTAT,MEDV)

set.seed(234)

trainIndex <- createDataPartition(boston$MEDV, 
                                  p = 2/3, 
                                  list = FALSE, 
                                  times = 1)

########### TRAIN /TEST SPLIT  ################
# removing CHAS since it is a factor and X matrix must be numeric
train <- data.frame(CHAS=boston[trainIndex,c(1)],
                    scale(boston[trainIndex,-c(1)]))
test <- data.frame(CHAS=boston[-trainIndex,c(1)],
                    scale(boston[-trainIndex,-c(1)]))

############## LINEAR REGRESSION (RIDGE)  ##############
#Cross-Validation of the lambda parameter 
y <- train$MEDV
x <- train[,-dim(train)[2]] %>% data.matrix()

lambda.max <- 1e5  
n.lambdas <- 50
lambdas <- exp(seq(10^-2,log(lambda.max+1),length=n.lambdas))-1 

mod <- glmnet::glmnet(x, y, alpha = 0, lambda = lambdas)

# specify alpha = 0 for ridge regression.
model.ridge <- glmnet::cv.glmnet(x, y, alpha = 0, lambda = lambdas)
#The lowest point in the curve indicates the optimal lambda: 
#the log value of lambda that best minimised the error in cross-validation.
lambda.ridge <- model.ridge$lambda.min

############## PLOT ############## 
library(plotmo)

glmcoef<-coef(mod,lambda.ridge )
coef.increase<-dimnames(glmcoef[glmcoef[,1]>0,0])[[1]]
coef.decrease<-dimnames(glmcoef[glmcoef[,1]<0,0])[[1]]

#get ordered list of variables as they appear at smallest lambda
allnames<-names(coef(mod)[,
            ncol(coef(mod))][order(coef(mod)[,
            ncol(coef(mod))],decreasing=TRUE)])

#remove intercept
allnames<-setdiff(allnames,allnames[grep("Intercept",allnames)])

#assign colors
cols<-rep("gray",length(allnames))
cols[allnames %in% coef.increase]<-"green"      # higher medv is good
cols[allnames %in% coef.decrease]<-"red"        # lower medv is not

plot_glmnet(mod,label=TRUE,s=lambda.ridge,col=cols)

```

To select the best model, we now use 10x10-CV using the lambda that best minimised 
the error in cross-validation, which is `r lambda.ridge`.

```{r, echo = FALSE, message=FALSE, warning=FALSE}

# ridge REGRESSION CV-10
## specify 10x10 CV
K <- 10; TIMES <- 10
trc <- trainControl (method="repeatedcv", number=K, repeats=TIMES)

model.ridge.10x10CV <- caret::train(MEDV ~ ., 
                                    data = train, 
                                    trControl=trc, 
                                    method='glmnet', 
                                    tuneGrid=expand.grid(alpha=0, lambda=lambda.ridge))

normalization.train <- (length(train$MEDV)-1)*var(train$MEDV)
NMSE.ridge.train.error <- crossprod(predict (model.ridge.10x10CV) - train$MEDV) / normalization.train


model.ridge.FINAL <- glmnet::glmnet(x, y, alpha = 0, lambda = lambda.ridge)

## This is the test NMSE for ridge:
normalization.test <- (length(test$MEDV)-1)*var(test$MEDV)
sse_raw <- test$MEDV - model.ridge.FINAL$a0- data.matrix(test[,-dim(train)[2]]) %*% model.ridge.FINAL$beta 
sse <- crossprod (as.matrix(sse_raw)) 

NMSE.ridge.test.error <- sse/normalization.test
```

So our final model has *Df=* `r model.ridge.FINAL$df` which is the number of non-zero coefficients 
and *%Dev=* `r model.ridge.FINAL$dev.ratio` is the percent deviance explained, which 
is quite good.

In terms of interpreting the coefficients, we observe that each additional room (*RM*) is associated
with an increase in the house price, on average. This is quite straightforward, in principle,
since it is to be expected that the larger the house, loosely speaking, the more expensive it 
will be. In addition, we see that an increase in *RAD* (index of accessibility to radial highways)
is associated with an increse in *MEDV*. So basically, if we were to think of the town 
as a graph we would be capturing the connectivity degree of a specific suburb; so a remote
node would have a lower value. Moreover, an increase in *CHAS* would mean that it will
take the value of 1 is associate with an increase in *MEDV*, so ultimately if the
Charles River passes through this suburb then this signals a higher  house price, on average.

On the other hand, an increase in *LSTAT* (% lower status of the population) is associated 
with a decrease in the house price, on average. Most interestinglty though is that the 
increase in *PTRATIO* (pupil-teacher ratio by town) is associated with a decrease in the 
house price, on average. So in other words, the education offering of a town increases its value. 
Also, an increase in *DIS* (weighted distances to five Boston employment centres)
is associated with a decrease in *MEDV*, on average. So, having to do a larger commute 
to work signals a lower house price. Another reasonable result is the fact that 
an increase in *NOX* (nitric oxides concentration) is associated with a decrease in *MEDV*,
so air pollution is a detractor to house price.

Furthermore, we see that neither *AGE* (proportion of owner-occupied units built prior to 1940) 
nor *INDUS* (roportion of non-retail business acres per town) seem to have a considerable effect on *MEDV*. 

Finally, we obtain the train and test error. 

```{r,  echo = FALSE, message=FALSE, warning=FALSE,results='asis'}
error.models <- tibble(regression_method = "ridge regression",
                        train_error = as.numeric(NMSE.ridge.train.error),
                        test_error = as.numeric(NMSE.ridge.test.error))



error.models%>%
  kable(format = "latex", booktabs = TRUE , digits = 3,
        caption = "Model Errors Summary") %>% 
  kable_styling(position = "center",latex_options = c("HOLD_position"),
                font_size = 9)
```

The difference between train and test errors is not that large, even if the test 
set is relatively small, and thus subject to a great variance.

