---
title: | 
  | \LARGE ASM Practice 
subtitle: "Local Linear Regression"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "20/11/2019"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(42)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, results="hide"}
requireorinstall=function(package=""){
  reqpac=parse(text=paste("require(",as.character(package),")"))
  if(eval(reqpac)){
    print(paste(as.character(package), "has been loaded correctly"))
  } else {
    print(paste("trying to install" ,as.character(package)))
    eval(parse(text=paste("try(install.packages(",as.character(package),"))")))
    if(eval(reqpac)){
      print(paste(as.character(package) ,"has been installed and loaded correctly"))
    } else {
      warning(paste("could not install",as.character(package)))
    }
  }
}
requireorinstall(c("sm","tidyverse","dplyr","KernSmooth","kable","kableExtra"))
```
The aim of this project is to compute the conditional variance $\sigma^2 (x)$ 
of the variable $lgWeight=log(Weight)$ of the *aircraft* dataset (in *sm*
package) given the year, *Yr*, variable.

```{r}
# Load data and pre-process
data("aircraft")
attach(aircraft)
lgPower <- log(Power)
lgSpan <- log(Span)
lgLength <- log(Length)
lgWeight <- log(Weight)
lgSpeed <- log(Speed)
lgRange <- log(Range)
```

## Estimation using *locpolreg* function

We load our local function *locpolreg* along with the *bandwith_selection* 
script which contains different functions for bandwidth selection.
We will choose the bandwidth hyper-parameter by LOOCV. We use the appropriate 
function in the *bandiwth_selection* script to get the LOOCV and GCV estimates
(which uses the *locpolreg* function).
Regarding the Kernel choice, we decide to use the *normal* kernel.

```{r}
# Load local functions
source("locpolreg.R")
source("bandwith_selection.R")
# Choice between "normal"  (Gaussian, default), 
# "epan"    (Epanechnikov) or 
# "rs.epan" (re-scaled Epanechnikov)
# "unif"    (Uniform Kernel in [-1,1])
kernel.type <- "normal"
# Define Bandwith candidates
h.v <- exp(seq(log(diff(range(Yr))/100), log(diff(range(Yr))/4),l=20))
# Dataframe for bandwith last comparsion table
h.df <- data.frame(Bandwith_Model_1=rep(0,2), Bandwith_Model_2=rep(0,2),
                     row.names = c("Locpolreg","Sm.Regression"))
# Dataframe for variance estimate last comparsion table
sigma.df <- data.frame(Locpolreg=rep(0,length(Yr)),
                       Sm.Regression=rep(0,length(Yr)))
```

The first thing we need to do is to compute the local linear regression for the predicted variable *lgWeight* depending on *Yr*. We choose the bandwidth as the one that minimizes the LOOCV estimate. After choosing the bandwidth, we build the local linear regression model using the *locpolreg* function and compute the residual values.

```{r, fig.align='center', out.height="170pt", fig.width=12}
op<-par(mfrow=c(1,2))

# Get LOOCV and GCV estimates
h.result <- h.cv.gcv(x=Yr, y=lgWeight, h.v=h.v, q=1, type.kernel=kernel.type) %>% 
  as.data.frame() %>% arrange(h.v)
# Plot LOOCV estimates vs values
plot(cv~h.v, h.result, type="b", main="Bandwith against LOOCV estimate",
     ylab="LOOCV estimate", xlab="Bandwith")
h.min <- h.result[which.min(h.result$cv),"h.v"]
h.df[1,1] <- h.min # Save bandwith for last table
abline(v=h.min, col=2)
text(x=h.min,y=h.result[which.min(h.result$cv),"cv"], labels
     =round(h.min,digits=2), pos = 3, col = 2)
# Local linear regresion model
m.result <- locpolreg(x=Yr,y=lgWeight,h=h.min,q=1,type.kernel=kernel.type,
                      main="Local Linear Regression model",
                      xlab="Yr",
                      ylab="lgWeight")
eps <- lgWeight - m.result$mtgr
eps2 <- eps*eps
Z <- log(eps2)

par(op)
```

Having obtained the residual values $\hat{\epsilon}_i = y_i - \hat{m}(x_i)$ we compute
their logarithm $z_i = \log{\hat{\epsilon}^2_i }$. We need to build a new model 
for $z_i$ against $x_i$.

```{r, fig.align='center', out.height="170pt", fig.width=12}
op<-par(mfrow=c(1,2))
h.v <- exp(seq(log(diff(range(Yr))/100), log(diff(range(Yr))/4),l=20))
# Get LOOCV and GCV estimates
h.result <- h.cv.gcv(x=Yr, y=Z, h.v=h.v, q=1, type.kernel=kernel.type) %>% 
  as.data.frame() %>% arrange(h.v)
plot(cv~h.v, h.result, type="b", main="Bandwith against LOOCV estimate",
     ylab="LOOCV estimate", xlab="Bandwith")
h.min <- h.result[which.min(h.result$cv),"h.v"]
h.df[1,2] <- h.min # Save bandwith for last table
abline(v=h.min, col=2)
text(x=h.min,y=h.result[which.min(h.result$cv),"cv"], labels
     =round(h.min,digits=2), pos = 3, col = 2)
# Build Local Linear Regression model
q.result <- locpolreg(x=Yr,y=Z,h=h.min,q=1,type.kernel=kernel.type,
                      main="Local Linear Regression model",
                      xlab="Yr",
                      ylab="lgWeight")
par(op)
```

Finally, the conditional variance $\hat{\sigma}^2(x) =\exp{\hat{q}(x)}$ where 
$\hat{q}(x)$ is the estimate of the previous model.

```{r}
sigma2 <- exp(q.result$mtgr)
sigma <- sqrt(sigma2)
sigma.df$Locpolreg <- sigma # Save variance estimate for last table
```

To sum up the results, we plot the value of $\epsilon_i^2$ against $x_i$ 
superimposing the values of $\hat{\sigma}^2(x)$, and also the values of 
$\hat{m}(x)$ with the bands $\hat{m}(x) \pm 1.96\hat{\sigma}(x)$.

```{r, fig.align='center', out.height="200pt", fig.width=12}
op<-par(mfrow=c(1,2))
plot(Yr, eps2, cex=0.8, main="Residuals against Yr with variance estimate",
     ylab="Squared Residuals")
points(Yr, sigma2, type="l", col="red", lwd=2)

plot(Yr,m.result$mtgr, main="Variance estimate with confidence bands",
     ylab="Variance estimate")
points(Yr,m.result$mtgr+1.96*sigma, type="l", col="blue", lty=2)
points(Yr,m.result$mtgr-1.96*sigma, type="l", col="blue", lty=2)
par(op)
```

## Estimation using *sm.regression* function

We use the function *dpill* in the *KernSmooth* package in order to compute the
*Plug-in* bandwidth parameter. Afterwards, we use this bandwidth with the
*sm.regression* function in the *sm* package to compute the local linear 
regression models, both $\hat{m}(x)$ and $\hat{q}(x)$.
Regarding the Kernel choice, we decide to use the *normal* kernel.

```{r, fig.align='center', out.width="250pt"}
h.min <- dpill(x=Yr, y=lgWeight)
h.df[2,1] <- h.min # Save bandwith for last table
```

After choosing the bandwidth, we build the local linear regression model using
the *sm.regression* function and compute the residual values $\hat{\epsilon}_i = y_i - \hat{m}(x_i)$ and their logarithm $z_i = \log{\hat{\epsilon}^2_i }$. We then build a new model for $z_i$ against $x_i$.

```{r, fig.align='center', out.height="200pt", fig.width=12}
op<-par(mfrow=c(1,2))

# Build local linear regression model
m.result <- sm.regression(x=Yr,y=lgWeight,h=h.min,
                          eval.grid=FALSE, eval.points=Yr)
title(main="lgWeights against Yr with Local Linear regression")
# Compute residuals
X <- m.result$eval.points # == Yr
eps <- lgWeight - m.result$estimate
eps2 <- eps*eps
Z <- log(eps2)
# Local linear regression model for residuals
h.min <- dpill(x=X, y=Z)
h.df[2,2] <- h.min # Save bandwith for last table
q.result <- sm.regression(x=X,y=Z,h=h.min,
                          eval.grid=FALSE, eval.points=X,
                          ylab="Log of Squared Residuals")
title(main="Log(Squared Residuals) vs Yr with Local Linear regression")

par(op)
```

Finally, the conditional variance $\hat{\sigma}^2(x) =\exp{\hat{q}(x)}$ where 
$\hat{q}(x)$ is the estimate of the previous model.

```{r}
sigma2 <- exp(q.result$estimate)
sigma <- sqrt(sigma2)
sigma.df$Sm.Regression <- sigma # Save variance estimate for last table
```

To sum up the results, we plot the value of $\epsilon_i^2$ against $x_i$ 
superimposing the values of $\hat{\sigma}^2(x)$, and also the values of 
$\hat{m}(x)$ with the bands $\hat{m}(x) \pm 1.96\hat{\sigma}(x)$.

```{r, fig.align='center', out.height="200pt", fig.width=12}
op<-par(mfrow=c(1,2))
plot(X, eps2, cex=0.8, main="Residuals against Yr with variance estimate",
     ylab="Squared Residuals")
points(X, sigma2, type="l", col="red", lwd=2)

plot(m.result$eval.points, m.result$estimate, main="Variance estimate with confidence bands",
     ylab="Variance estimate", xlab="Yr")
points(m.result$eval.points, m.result$estimate+1.96*sigma,
       type="l", col="blue", lty=2)
points(m.result$eval.points, m.result$estimate-1.96*sigma,
       type="l", col="blue", lty=2)
par(op)
```

## Method comparison

To compare the results of the methods, we present a table with the chosen bandwiths for both models:

```{r,  echo = FALSE, message=FALSE, warning=FALSE,results='asis'}
h.df %>%
  kable(format = "latex", booktabs = TRUE , digits = 3,
        caption = "Model Bandwith comparison") %>% 
  kable_styling(position = "center",latex_options = c("HOLD_position"),
                font_size = 9)
```
```{r, fig.align='center', out.width="400pt"}
plot(Yr, sigma.df$Locpolreg, col=2, pch=18,
     ylim = c(min(min(sigma.df$Locpolreg),min(sigma.df$Sm.Regression)),
              max(max(sigma.df$Locpolreg),max(sigma.df$Sm.Regression))),
     main="Variance estimate comparison between models",
     xlab="Yr", ylab="Variance estimate")
points(Yr, sigma.df$Sm.Regression, col=3,  pch=20)
legend("topleft", legend = c("LocPolReg","Sm.Regression"), col=2:3,
       pch=c(18,20))
```

The bandwith values do not differ greatly. In addition, by looking at the comparison plot, we confirm the previous idea. However, we observe a slightly more extreme behaviour for the *Sm.Regression* estimate, due to its lower bandiwth (for model 2).
