---
title: | 
  | \LARGE ASM Practice 
subtitle: "Local Poisson Regression"
author: "Maria Gkotsopoulou & Ricard Monge Calvo & Amalia Vradi"
date: "14/12/2019"
geometry: margin=1.5cm
output: 
  pdf_document: 
    latex_engine: xelatex
fontsize: 11pt
spacing: single
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \usepackage{subfig}
  \titlespacing{\section}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{10pt plus 1pt minus 1pt}{0pt plus 1pt minus 1pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(42)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, results="hide"}
requireorinstall=function(package=""){
  reqpac=parse(text=paste("require(",as.character(package),")"))
  if(eval(reqpac)){
    print(paste(as.character(package), "has been loaded correctly"))
  } else {
    print(paste("trying to install" ,as.character(package)))
    eval(parse(text=paste("try(install.packages(",as.character(package),"))")))
    if(eval(reqpac)){
      print(paste(as.character(package) ,"has been installed and loaded correctly"))
    } else {
      warning(paste("could not install",as.character(package)))
    }
  }
}
requireorinstall(c("sm","tidyverse","dplyr","KernSmooth","kable","kableExtra"))
```
The aim of this project is to obtain a function to choose the bandwith 
hyperparameter for local Poisson regression by using LOOCV estimates and apply
it to the Country Development dataset [^1].

[^1]: [Human Development Data (1990-2017)](http://hdr.undp.org/en/data).

This file contains the following variables:

* `Life.expec` Life expectancy at birth.
* `Life.expec.f` Life expectancy at birth for females.
* `Life.expec.m` Life expectancy at birth for males.
* `le.fm` Difference `Life.expec.f` minus `Life.expec.m`.
* `Inf.Mort.rat` Infant mortality rate: The annual number of deaths of infants under one year of age per 1,000 live births in the same year.
* `Agric.employ.%` Employment in agriculture (% of total employment).

We see the variable *le.fm* about life expectancy difference between male and
female is always positive and can be rounded to give a Poisson-like 
distribution.

```{r, echo=FALSE, warning=FALSE, out.width='300pt', fig.align='center'}
countries<-read.csv2(file="HDI.2017.subset.csv",row.names = 1)
attach(countries)
le.fm.r <- round(le.fm)
hist(le.fm.r, br=10,
       main="Histogram on life expectancy difference (rounded)")
```

## Bandwith choice functions for local Poisson regression

We modify the functions *h.cv.sm.binomial* and *loglik.CV* to obtain a bandwidth
choice method for the local Poisson regression, based on the LOOCV estimation of
the expected log-likelihood of an independent observation.

In the fuction *loglik.CV*, the process of fitting a *sm.poisson* model 
considering all datapoints apart from one is repeated $n$ times. From each fitted 
model we obtain the estimates $\lambda_i$ used to compute the log-likelihood.
We use the following expression:

$$
l_{CV} = \frac{1}{n} \sum_{i=1}^{n} \log{\left( e^{-\lambda_i} \frac{\lambda_i^{y_i}}{y_i!}  \right)} = 
 \frac{1}{n} \sum_{i=1}^{n} \left(-\lambda_i + y_i \log{\lambda_i} 
 - \log{y_i!} \right)
$$
where $n$ is the number of instances in the dataset, $y_i$ is each of the 
response values and the expression $e^{-\lambda_i} \frac{\lambda_i^{y_i}}{y_i!}$
corresponds to the estimate of $Pr(Y=y_i|X=x_i)$, the probability of getting the
response value given the predictors $x_i$ and the estimate of the distributions parameter $\lambda_i$.

On the other hand, we rename the function *h.cv.sm.binomial* to 
*h.cv.sm.poisson* and change the selected bandwidth parameter *h.cv* as the one 
that maximizes the LOOCV estimates of the expected log-likelihood returned by 
the method function (*loglik.CV* in our case).

```{r, echo=TRUE}
# Bandwidth choice in the local Poisson regression 
# by leave-one-out cross-validation.
# Function "sm.poisson", from library "sm", is used.
# method can be equal to 'loglik.CV' or other method which needs to be MAXIMIZED
h.cv.sm.poisson <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV){
   cv.h <- numeric(l.h)
   if (is.null(rg.h)){
      hh <- c(h.select(x,y,method="cv"),
              h.select(x,y,method="aicc"))
      rg.h <- range(hh)*c(1/1.1, 1.5)
   }
   i <- 0
   gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
   for (h in gr.h){
      i <- i+1
      cv.h[i] <- method(x,y,h)
   }
   return(list(h = gr.h, 
               cv.h = cv.h, 
               h.cv = gr.h[which.max(cv.h)])) # Maximizing log-likelihood
}
# method loglik.CV: leave-one-out log-likelihood 
loglik.CV <- function(x,y,h){
  n <- length(x)
  pred <- sapply(1:n, 
      function(i,x,y,h){
         sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
      },   x,y,h)
  return(sum(y*log(pred) - pred - log(factorial(y)))/n)
}
```

## Local Poisson regression for Country Development Data



We want to build a local Poisson regression to model life expectancy differnece between male and female *le.fm.r*, rounded to integer values, as a function of *Life.expec*, for the Country Development data. We use the previous function `h.cv.sm.poisson` to choose the bandwidth hyperparameter.

```{r, out.width='350pt', fig.align='center'}
# Using h.CV.loglik
h.CV.loglik <- h.cv.sm.poisson(Life.expec, le.fm.r,
                               rg.h=c(0.5,20),method=loglik.CV)
{
  plot(h.CV.loglik$h, h.CV.loglik$cv.h, col="purple", type="b",
       main="LOOCV estimation of the expected Log-likelihood vs Bandwidth",
       ylab = "Expected Log-likelihood", xlab = "Bandwidth")
  abline(v=h.CV.loglik$h.cv, col="blue")
  text(h.CV.loglik$h.cv, min(h.CV.loglik$cv.h),
       labels = paste0("h.cv.loglik =", round(h.CV.loglik$h.cv,2)),
       col="blue", pos=4)
}
```

Assuming that the target variable, given the predictor, follows a Poisson
distribution, our model estimates its mean rate $\lambda$. We show the
scatterplot of the target-predictor points together with our model's estimated
rates. In addition, we fit a Generalized Linear Model of the Poisson family, 
using the *glm* function, and plot its estimated rates for comparison purposes.

```{r, warning=FALSE, message=FALSE, out.width='350pt', fig.align='center'}
{
  # Local Poisson regresion
  aux <- sm.poisson(Life.expec, le.fm.r, h=h.CV.loglik$h.cv, col=1)
  title(main="Scatterplot of target-predictor points with estimated parameters")
  # Generalized Poisson regresion
  aux.glm <- glm(le.fm.r ~ Life.expec, family=poisson)
  pred <- predict(aux.glm, 
                  newdata=data.frame(Life.expec=aux$eval.points),
                  type="response")
  lines(aux$eval.points, pred, col=2, lty=2)
  legend("topleft",c("Loc.Pois.reg","GLM.Pois"),
        col=c(1,2),lty=c(1,2))
}
```

Estimates are similar between both models but the local Poisson regression seems
to better fit to our sample.